Turbulence and Particle Acceleration in Giant Radio Haloes: the Origin of Seed Electrons

### 2016-06-19 paper.tex
notation unified throughout the paper, text edited

\title[Origin of Seed Electrons]{Turbulence and Particle Acceleration in Giant Radio Halos: the Origin of Seed Electrons}

@@

\begin{abstract}
  About one third of X-ray-luminous clusters show smooth, unpolarized
  radio emission on $\sim$Mpc scales, known as giant radio halos. One
  promising model for radio halos is Fermi II acceleration of seed
  relativistic electrons by compressible turbulence in the
  intracluster medium (ICM). The origin of these seed electrons has never
  been fully explored. Here, we integrate the Fokker-Planck equation
  of the cosmic ray (CR) electron and proton distributions in
  cosmological simulations of cluster formation, and confront them with radio surface brightness and spectral observations of Coma. For standard
  assumptions, structure formation shocks lead to a seed electron
  population which produces too centrally concentrated radio
  emission. By modifying properties of the CR population (rapid streaming; enhanced CR electron acceleration at shocks) or turbulence (increasing $\eps_{\rm turb}/\eps_{\rm therm}$ with radius), we can match observations, but only at the expense of fine-tuning. We then perform a parameter study. Under naive assumptions, radio properties are exponentially sensitive to the amplitude of turbulence, which is inconsistent with small scatter in scaling relations. This sensitivity can be removed if we relate the acceleration time to the turbulent dissipation time. Their ratio depends only on plasma parameters. In this case, turbulence above a threshold value required to overcome cooling provides a fixed amount of amplification; observations could then potentially constrain the unknown CR seed population. Understanding the small scatter in radio halo scaling relations may prove a rich source of insight on plasma processes in the ICM.   
\end{abstract}

@@

% --- section: Introduction --- %
\section{Introduction}
About one third of X-ray-luminous clusters show smooth, unpolarized
radio emission on $\sim$Mpc scales, known as giant radio halos (RHs)
\citep{2014IJMPD..2330007B}. They appear only in disturbed, merging
clusters and the RH luminosity correlates with the X-ray luminosity
\citep{2001A&A...369..441G,2012A&ARv..20...54F} and the Compton
$y$-parameter \citep{2012MNRAS.421L.112B,2013A&A...554A.140P}. The RHs
show that CR electrons and magnetic fields permeate a large volume
fraction of the intra-cluster medium (ICM). The dominant CR source,
given the smoothness and enormous extent of RHs, is thought to be
structure formation shocks \citep{miniati01,pfrommer08}. At the same
time, plasma processes, the origin of magnetic fields and particle
acceleration in a turbulent, high-$\beta$ plasma like the ICM are not
well understood. Radio halos thus provide an incisive probe of
non-thermal processes in the ICM.

@@

The alternative model for RHs is re-energization of seed suprathermal
electrons by Fermi II acceleration when ICM turbulence becomes
transonic during mergers
\citep{1987A&A...182...21S,1993ApJ...406..399G,2001MNRAS.320..365B,
  2004MNRAS.350.1174B,brunetti07,brunetti11,miniati15}. Due
to the short radiative cooling time of high-energy relativistic
electrons, the cluster synchrotron emission quickly fades away after a
merger, which naturally explains the observed bimodality of RHs
\cite[see e.g.][]{2013MNRAS.429.3564D,2014MNRAS.443.3564D}.

@@

While analytic estimates have been made, there has been no ab initio
demonstration that structure formation can lead to the required
abundance of seed electrons with the correct spatial and spectral
characteristics. This is a non-trivial requirement: Coulomb cooling in
dense cluster cores is severe, and DSA fossil electrons may not
survive. On the other hand, for secondaries to constitute the seed
population, the CRp population required in the best-studied case of
the Coma cluster must have a very broad and flat (or even slightly
inverted) spatial profile \citep{brunetti12}, in contrast with the
thermal plasma whose energy density declines steeply with radius. In
Fig.~\ref{fig:Edens} we show that such a distribution is not predicted
by cosmological simulations \cite[see
  also][]{pinzke10,2014MNRAS.439.2662V}. If CRps are predominantly
advected with the cluster plasma, their distribution will be peaked
towards the cluster center and shows a similar characteristics as the
thermal plasma. As a consequence, the distribution of secondary
electrons and the resulting radio synchrotron emission is also peaked
since the hadronic reaction is a two-body scattering process. Hence,
the simulated emission falls short of the observed extended and flat
radio profile of the Coma cluster.

\begin{figure}
  \includegraphics[width=1.0\columnwidth]{fCR.radius.coma.g72a.Rad14.2400p.z0.NL.xKR.eb23.eI067.DII.140.v6.eps}
  \caption{Spatial distribution of CRp energy density in the Coma
    cluster. The red dash-dotted line shows the required distribution
    of seed CRps that generate secondary electrons via proton-proton
    (p-p) collisions required to reproduce Coma radio brightness
    observations after Fermi-II reacceleration \citep{brunetti12}. The
    blue solid line shows the distribution of fossil CRps found in
    cosmological simulations, which disagrees with the required
    profile. To better compare the two models in this figure, we
    normalize the required distribution of CRps by fixing the total
    CRp energy $E_\rmn{CR}$ to 0.3 percent of the total thermal
    energy, consistent with observations
    \citep{2014ApJ...787...18A,2012ApJ...757..123A}.}
  \label{fig:Edens}
\end{figure}

Indeed, arriving at a seed population with the required
characteristics is highly constraining, and has the potential to teach
us much about the origin of CRps/CRes in clusters.  In this work, we
use our hydrodynamical zoom simulations of galaxy clusters in a
cosmological setting to follow the distribution functions of seed
populations for CRps and CRes, and integrate the Fokker-Planck
equation of CR transport along Lagrangian particle trajectories. We
model diffusive shock acceleration at structure formation shocks,
and account for various loss processes of CRs. Utilizing new insights from our recent work on DSA generated
  fossil electrons \citep{pinzke13}, we generate the first
  quantitative calculation of primary and secondary seed electrons

@@

In this paper, we explore how the radio surface brightness profile and spectrum of the best known radio halo, Coma, can be used to constrain the underlying properties of the seed cosmic rays and turbulence. We aim to constrain the normalization and spatial profile of these two input ingredients in turbulent reacceleration models. The outline of this paper is as follows. In \S\ref{sect:method}, we outline the basic physics of turbulence reacceleration of cosmic rays which we use. In \S\ref{sec:results}, we use cosmological simulations to generate a seed CR population, and combine it with our parameterized model of turbulence to produce radio surface brightness profiles and spectra of Coma. We find that it is possible to fit the observations using physically motivated modifications of the seed population    
(rapid streaming; enhanced CR electron acceleration at shocks) or turbulence (increasing $\eps_{\rm turb}/\eps_{\rm therm}$ with radius), but only at the expense of fine tuning. In \S\ref{sect:param_comp}, we explore the reason for this fine-tuning, and seek ways to overcome it. We perform a parameter study on spherically symmetric, static models where we vary properties of the seed population and turbulence. We find exponential sensitivity to the amount of turbulent, which can be eliminated if the turbulent acceleration and dissipation time are linked. In this case, turbulence above a threshold value required to overcome cooling provides a fixed amount of amplification; observations could then potentially constrain the unknown CR seed population. We summarize and conclude in \S\ref{sec:conclusions}.

@@

\subsection{Basic equations}
As previously noted, secondaries produced by shock accelerated CRp
have the wrong spatial profile to explain RH observations. Because
they arise from a two body process, they are too centrally
concentrated. They also produce gamma-ray emission in excess of
Fermi-LAT upper limits
\citep{2012ApJ...757..123A,brunetti12,2014ApJ...787...18A}.

Given a seed population of CRs, we adopt essentially the same set of
plasma physics assumptions as the reacceleration model for RHs
\citep{brunetti07,brunetti11}. We solve the isotropic, gyro-phase
averaged Fokker-Planck equation (via a Crank-Nicholson scheme) for the
time evolution of the CRe distribution in the Lagrangian frame
\citep{brunetti07,brunetti11}:
\begin{eqnarray}
{{d f_{\rmn{e}}(p,t)}\over{d t}} &\!=&
\frac{\partial}{\partial p}
\left\{
f_{\rmn{e}}(p,t)\left[
\left|\frac{dp}{dt}\right|_{\rm Coul} 
+ \frac{p}{3}\left(\bnabla\bcdot \bvel\right)
+ \left|\frac{dp}{dt}\right|_{\rm rad}\right.\right.
\nonumber\\
&-& \left.\left.{1\over{p^2}}{{\partial }\over{\partial p}}\left(p^2 D_{\rm pp}\right) 
\right]\right\} - \left(\bnabla\bcdot \bvel\right) f_{\rmn{e}}(p,t)
\nonumber\\
&+& {{\partial^2 }\over{\partial p^2}}
\left[
D_{\rm pp} f_{\rmn{e}}(p,t) \right]+ Q_{\rmn{e}}\left[p,t;f_{\rmn{p}}(p,t)\right]   \, .
\label{elettroni}
\end{eqnarray}
Here $f_{\rmn{e}}$ is the one-dimensional distribution in position $x$
(suppressed for clarity), momentum $p$ and time $t$ (which is
normalized such that the number density is given by
$n_{\rmn{e}}(t)=\int d p f_{\rmn{e}}(p,t)$), $d/dt=\partial/\partial
t+\bvel\bcdot\bnabla$ is the Lagrangian derivative, $\bvel$ is the gas
velocity, $|dp/dt|$ represents Coulomb
\citep[Coul,][]{1972Phy....60..145G} and radiative
\citep[rad,][]{1979rpa..book.....R} losses, respectively,
\begin{eqnarray}
   \left|\frac{dp}{dt}\right|_{\rm Coul} &=& \frac{3\,\sigma_\rmn{T}\,n_e\,c}{2\, \beta^2}
  \left[\ln\left(\frac{m_e c^2 \beta \sqrt{\gamma-1}}{\hbar\,\omega_\rmn{plasma}}\right)\right.\nonumber \\
    &-&\ln(2)\left(\frac{\beta^2}{2}+\frac{1}{\gamma}\right)+\frac{1}{2}+
    \left.\left(\frac{\gamma-1}{4\gamma}\right)^2\right]\,, \\
  \left|\frac{dp}{dt}\right|_{\rm rad~} &=& \frac{4}{3}\frac{\sigma_\rmn{T}}{m_ec}\,
  \frac{p^2}{\beta}\,\left[1+\left(\frac{B}{B_\rmn{CMB}}\right)^2\right]\,.
\end{eqnarray}
Here $\beta = p/\sqrt{1+p^2}$ is the dimensionless velocity of CRs,
$\gamma=\sqrt{1+p^2}$ is the Lorentz factor of CRs,
$\omega_\rmn{plasma} = \sqrt{4\pi e^2 n_e / m_e}$ is the plasma
frequency, $n_e$ is the number density of free electrons, and
$\sigma_\rmn{T}= 8\pi e^4/3(m_e c^2)^2$ is the Thomson cross
section. The {\it rms} magnetic field strength is denoted by $B$ and
the equivalent field strength of the cosmic-microwave background is
given by $B_\rmn{CMB} = 3.24 (1 + z)^2\mu\rmn{G}$, where $z$ denotes
the redshift. In the peripheral cluster regions, where $B \ll
B_\rmn{CMB}$, the CRes loose virtually all their energy by means of
inverse Compton emission. $D_{\rm pp}$ is the momentum space diffusion
coefficient (see \S\ref{sec:reacc}), and
$Q_{\rmn{e}}$ denotes the injection rate of primary and secondary
electrons in the ICM (see Sect.~\ref{sec:cosmo_sim}). The first term
containing the expression $\bnabla\bcdot \bvel$ represents Fermi-I
acceleration and the second term of this form describes adiabatic
gains and losses.

During post-processing of our Coma-like cluster simulation, we solve
the Fokker-Planck equation over a redshift interval from $z=5$ to
0. The simulated cluster undergoes a major merger over the last
1-2~Gyrs that is thought to inject large turbulent eddies. As is commonly assumed \citep{brunetti07,brunetti11,2004ApJ...614..757Y,2013ApJ...771..131B}
we assume that about one Gyr after core passage the fields have
decayed down to the smallest scales $k_{\rm cut}$, and the radio halo turns on
shortly after. We choose this simulation snapshot to analyze. We are not very sensitive to this assumption, since the thermal and CR quantities are very similar a few
100 Myrs before and after z = 0, where we have chosen to evaluate the
simulations. In all our
calculations we assume that turbulent reacceleration efficiently
accelerates particles for $\tau{\rm cl} \sim 650$ Myrs (which is roughly the cascade
time on which turbulence is damped) and that during this turbulent
phase CR streaming and spatial diffusion can be neglected. In \S\ref{sect:self-limiting}, we explore sensitivity to the last assumption.

Thus far, we have ignore CR transport. However, if CRps stream in the ICM, then their spatial profile could
potentially flatten sufficiently \citep{ensslin11,wiener13}. This
scenario is very attractive: it generates seed electrons with the
right spatial footprint, and by removing CRps from the core, obeys
gamma-ray constraints. Turbulence plays two opposing roles:
Alfv{\'e}nic turbulence damps waves generated by the CR streaming
instability \citep{yan02,farmer04}, thus reducing self-confinement;
but compressible fast modes scatter CRs directly. Turbulent damping is
still efficient for highly subsonic conditions \citep{wiener13}, while
we assume compressible fast modes to only provide effective spatial
confinement during the periods of transonic, highly super-Alfv{\'e}nic
($M_{\rm A} \sim 5$) turbulence associated with mergers. Thus, CRs can
stream out when the cluster is kinematically quiescent. Furthermore,
even Alfv{\'e}nic streaming timescales are relatively short
\cite[$\sim 0.1-0.5$ Gyr;][]{wiener13} compared to the timescale on
which the CRp population is built up. Based on these findings, we
adopt a toy model for our \Mstream scenario in which CR streaming
instantaneously produces flat CRp profiles. We assume that CRs cannot stream significantly past
perpendicular $B$-fields at the accretion shock, so that the total
number of CRs is conserved within the virial radius during the
streaming process.
In all our calculations we assume that turbulent reacceleration
efficiently accelerates particles for $\tau_\rmn{cl}\sim650$~Myrs
(which is roughly the cascade time on which turbulence is damped) and
that during this turbulent phase CR streaming and spatial diffusion
can be neglected.

The time evolution of the spectral energy distribution of CRps,
$f_{\rmn{p}}$, is similarly given by:
\begin{eqnarray}
\lefteqn{
  {{d f_{\rmn{p}}(p,t)}\over{d t}} =
  {{\partial }\over{\partial p}}
  \left\{
  f_{\rmn{p}}(p,t)\left[ \left|{{dp}\over{dt}}\right|_{\rm Coul}
    + \frac{p}{3}\bnabla\bcdot \left(\bvel+\bvel_{\rmn{st}}\right)\right.\right.}
\nonumber\\
&&-\left.\left. {1\over{p^2}}{{\partial }\over{\partial p}}\left(p^2 D_{\rm pp}\right)
\right]\right\} - f_{\rmn{p}}(p,t) \bnabla\bcdot \left(\bvel+\bvel_{\rmn{st}}\right) 
- \bvel_{\rmn{st}}\bcdot\bnabla f_{\rmn{p}}(p,t)
\nonumber\\
&&+\,\,\, {{\partial^2 }\over{\partial p^2}}
\left[ D_{\rm pp} f_{\rmn{p}}(p,t) \right] - {{f_{\rmn{p}}(p,t)}\over{\tau_{\rm had}(p)}}
+ Q_{\rmn{p}}(p,t)\, ,
\label{eq:FP_p}
\end{eqnarray}
where $\bvel_{\rmn{st}}=-\varv_{\rmn{A}}\bnabla f_{\rmn{p}}/|\bnabla
f_{\rmn{p}}|$ is the streaming velocity in the isotropic transport
approximation, $\varv_{\rmn{A}}$ is the Alfv\'en speed,
and $Q_{\rmn{p}}(p,t)$ denotes the injection rate of shock accelerated
CRps as a function of momentum $p$ and time $t$ (see
Sect.~\ref{sec:cosmo_sim}). The timescale of
hadronic losses that produce pions via CRp collisions with thermal
protons of the ICM is given by 
\begin{eqnarray}
  \tau_{\rm had} = \left[c\,n_\rmn{th}\,\sigma^{+/-,0}(p)\right]^{-1}\,,
\end{eqnarray}
where we use the cross-section, $\sigma^{+/-,0}(p)$, given by the
fitting formula in \citep{1986ApJ...307...47D}.

\subsection{Turbulent Reacceleration}
\label{sec:reacc}

In turbulent reacceleration, particles gain energy via Fermi II acceleration. Wave-particle energy exchange takes place via transit time damping \citep{brunetti07,brunetti11}. The
TTD resonance requires the wave frequency to obey the resonance
condition, $\omega=k_\parallel\varv_\parallel$, where $k_\parallel$
and $\varv_\parallel$ are the parallel (projected along the magnetic
field) wavenumber of the compressible mode and particle velocity, respectively. This implies
that the particle transit time across the confining wave region
matches the wave period,
$\lambda_{\parallel}/\varv_{\parallel}=\tau_{\rmn{wave}}$ (hence the moniker, 'transit time damping'). Note that the CRs'
gyroradius does not enter the resonance condition. Hence the CRs that
are in resonance with compressible waves experience Fermi-II
acceleration irrespectively of the length scale of the perturbation.

However, the resonance changes the component of the particle momentum
parallel to the mean magnetic field, which over time leads to
increasing anisotropy in the particle distribution that decreases the
efficiency of reacceleration with time. As in \citet{brunetti11}, we
assume that there exists a mechanism---such as the firehose
instability---that isotropizes the CR distribution function at the
gyroscale and on the reacceleration time scale, which ensures
sustained efficient reacceleration with time.

>>>>

\title[Origin of Seed Electrons]{Turbulence and Particle Acceleration in Giant Radio Haloes: the Origin of Seed Electrons}

@@

\begin{abstract}
  About one third of X-ray-luminous clusters show smooth, unpolarised radio
  emission on $\sim$Mpc scales, known as giant radio haloes. One promising model
  for radio haloes is Fermi II acceleration of seed relativistic electrons by
  compressible turbulence in the intracluster medium (ICM). The origin of these
  seed electrons has never been fully explored. Here, we integrate the
  Fokker-Planck equation of the cosmic ray (CR) electron and proton
  distributions in cosmological simulations of cluster formation, and confront
  them with radio surface brightness and spectral observations of Coma. For
  standard assumptions, structure formation shocks lead to a seed electron
  population which produces too centrally concentrated radio emission. By
  modifying properties of the CR population (rapid streaming; enhanced CR
  electron acceleration at shocks) or turbulence (increasing
  turbulent-to-thermal energy density with radius), we can match observations,
  but only at the expense of fine-tuning. We then perform a parameter
  study. Under naive assumptions, radio properties are exponentially sensitive
  to the amplitude of turbulence, which is inconsistent with small scatter in
  scaling relations. This sensitivity can be removed if we relate the
  acceleration time to the turbulent dissipation time. Their ratio depends only
  on plasma parameters. In this case, turbulence above a threshold value
  required to overcome cooling provides a fixed amount of amplification;
  observations could then potentially constrain the unknown CR seed
  population. Understanding the small scatter in radio halo scaling relations
  may prove a rich source of insight on plasma processes in the ICM.
\end{abstract}

@@

% --- section: Introduction --- %
\section{Introduction}
About one third of X-ray-luminous clusters show smooth, unpolarised radio
emission on $\sim$Mpc scales, known as giant radio haloes (RHs)
\citep{2014IJMPD..2330007B}. They appear only in disturbed, merging clusters and
the RH luminosity correlates with the X-ray luminosity
\citep{2001A&A...369..441G,2012A&ARv..20...54F} and the Compton $y$-parameter
\citep{2012MNRAS.421L.112B,2013A&A...554A.140P}. The RHs show that CR electrons
and magnetic fields permeate a large volume fraction of the intra-cluster medium
(ICM). The dominant CR source, given the smoothness and enormous extent of RHs,
is thought to be structure formation shocks \citep{miniati01,pfrommer08}. At the
same time, plasma processes, the origin of magnetic fields and particle
acceleration in a turbulent, high-$\beta$ plasma (in which the thermal pressure
predominates the magnetic pressure) like the ICM are not well understood. Radio
haloes thus provide an incisive probe of non-thermal processes in the ICM.

@@

The alternative model for RHs is re-energisation of seed suprathermal
electrons by Fermi II acceleration when ICM turbulence becomes
transsonic during mergers
\citep{1987A&A...182...21S,1993ApJ...406..399G,2001MNRAS.320..365B,
  2004MNRAS.350.1174B,brunetti07,brunetti11,miniati15}. Due
to the short radiative cooling time of high-energy relativistic
electrons, the cluster synchrotron emission quickly fades away after a
merger, which naturally explains the observed bimodality of RHs
\cite[see e.g.][]{2013MNRAS.429.3564D,2014MNRAS.443.3564D}.

@@

While analytic estimates have been made, there has been no ab initio
demonstration that structure formation can lead to the required
abundance of seed electrons with the correct spatial and spectral
characteristics. This is a non-trivial requirement: Coulomb cooling in
dense cluster cores is severe, and DSA fossil electrons may not
survive. On the other hand, for secondaries to constitute the seed
population, the CRp population required in the best-studied case of
the Coma cluster must have a very broad and flat (or even slightly
inverted) spatial profile \citep{brunetti12}, in contrast with the
thermal plasma whose energy density declines steeply with radius. In
Fig.~\ref{fig:Edens} we show that such a distribution is not predicted
by cosmological simulations \cite[see
  also][]{pinzke10,2014MNRAS.439.2662V}. If CRps are predominantly
advected with the cluster plasma, their distribution will be peaked
towards the cluster centre and shows a similar characteristics as the
thermal plasma. As a consequence, the distribution of secondary
electrons and the resulting radio synchrotron emission is also peaked
since the hadronic reaction is a two-body scattering process. Hence,
the simulated emission falls short of the observed extended and flat
radio profile of the Coma cluster.

\begin{figure}
  \includegraphics[width=1.0\columnwidth]{fCR.radius.coma.g72a.Rad14.2400p.z0.NL.xKR.eb23.eI067.DII.140.v6.eps}
  \caption{Spatial distribution of CRp energy density in the Coma
    cluster. The red dash-dotted line shows the required distribution
    of seed CRps that generate secondary electrons via proton-proton
    (p-p) collisions required to reproduce Coma radio brightness
    observations after Fermi-II reacceleration \citep{brunetti12}. The
    blue solid line shows the distribution of fossil CRps found in
    cosmological simulations, which disagrees with the required
    profile. To better compare the two models in this figure, we
    normalise the required distribution of CRps by fixing the total
    CRp energy $E_\rmn{CR}$ to 0.3 per cent of the total thermal
    energy, consistent with observations
    \citep{2014ApJ...787...18A,2012ApJ...757..123A}.}
  \label{fig:Edens}
\end{figure}

Indeed, arriving at a seed population with the required characteristics is
highly constraining, and has the potential to teach us much about the origin of
CRps/CRes in clusters.  In this work, we use our hydrodynamical zoom simulations
of galaxy clusters in a cosmological setting to follow the distribution
functions of seed populations for CRps and CRes, and integrate the Fokker-Planck
equation of CR transport along Lagrangian particle trajectories. We model
diffusive shock acceleration at structure formation shocks, and account for
various loss processes of CRs. Utilising new insights from our recent work on
DSA generated fossil electrons \citep{pinzke13}, we generate the first
quantitative calculation of primary and secondary seed electrons

@@

In this paper, we explore how the radio surface brightness profile and spectrum
of the best known radio halo, Coma, can be used to constrain the underlying
properties of the seed CRs and turbulence. We aim to constrain the normalisation
and spatial profile of these two input ingredients in turbulent reacceleration
models. The outline of this paper is as follows. In \S\ref{sect:method}, we
outline the basic physics of turbulent reacceleration of CRs which we use. In
\S\ref{sec:results}, we use cosmological simulations to generate a seed CR
population, and combine it with our parametrized model of turbulence to produce
radio surface brightness profiles and spectra of Coma. We find that it is
possible to fit the observations using physically motivated modifications of the
seed population (rapid streaming; enhanced CR electron acceleration at shocks)
or turbulence (increasing the turbulent-to-thermal energy density, $\eps_{\rm
  turb}/\eps_{\rm th}$ with radius), but only at the expense of fine
tuning. In \S\ref{sect:param_comp}, we explore the reason for this fine-tuning,
and seek ways to overcome it. We perform a parameter study on spherically
symmetric, static models where we vary properties of the seed population and
turbulence. We find exponential sensitivity to the amount of turbulence, which
can be eliminated if the turbulent acceleration and dissipation time are
linked. In this case, turbulence above a threshold value required to overcome
cooling provides a fixed amount of amplification; observations could then
potentially constrain the unknown CR seed population. We summarise and conclude
in \S\ref{sec:conclusions}.

@@

\subsection{Basic equations}
As previously noted, secondaries produced by shock accelerated CRp
have the wrong spatial profile to explain RH observations. Because
they arise from a two body process, they are too centrally
concentrated. They also produce gamma-ray emission in excess of
Fermi-LAT upper limits
\citep{2012ApJ...757..123A,brunetti12,2014ApJ...787...18A,Ahnen2016}.

Given a seed population of CRs, we adopt essentially the same set of
plasma physics assumptions as the reacceleration model for RHs
\citep{brunetti07,brunetti11}. We solve the isotropic, gyro-phase
averaged Fokker-Planck equation (via a Crank-Nicholson scheme) for the
time evolution of the CRe distribution in the Lagrangian frame
\citep{brunetti07,brunetti11}:
\begin{eqnarray}
{{d f_{\rmn{e}}(p,t)}\over{d t}} &\!=&
\frac{\partial}{\partial p}
\left\{
f_{\rmn{e}}(p,t)\left[
\left|\frac{d p}{d t}\right|_{\rm Coul} 
+ \frac{p}{3}\left(\bnabla\bcdot \bvel\right)
+ \left|\frac{d p}{d t}\right|_{\rm rad}\right.\right.
\nonumber\\
&-& \left.\left.{1\over{p^2}}{{\partial }\over{\partial p}}\left(p^2 D_{pp}\right) 
\right]\right\} - \left(\bnabla\bcdot \bvel\right) f_{\rmn{e}}(p,t)
\nonumber\\
&+& {{\partial^2 }\over{\partial p^2}}
\left[
D_{pp} f_{\rmn{e}}(p,t) \right]+ Q_{\rmn{e}}\left[p,t;f_{\rmn{p}}(p,t)\right]   \, .
\label{elettroni}
\end{eqnarray}
Here $f_{\rmn{e}}$ is the one-dimensional distribution in position $x$
(suppressed for clarity), momentum $p$ and time $t$ (which is
normalised such that the number density is given by
$n_{\rmn{e}}(t)=\int d p f_{\rmn{e}}(p,t)$), $d/d t=\partial/\partial
t+\bvel\bcdot\bnabla$ is the Lagrangian derivative, $\bvel$ is the gas
velocity, $|d p/d t|$ represents Coulomb
\citep[Coul,][]{1972Phy....60..145G} and radiative
\citep[rad,][]{1979rpa..book.....R} losses, respectively,
\begin{eqnarray}
   \left|\frac{d p}{d t}\right|_{\rm Coul} &=& \frac{3\,\sigma_\rmn{T}\,n_{\rmn{e}}\,c}{2\, \beta_{\rmn{e}}^2}
  \left[\ln\left(\frac{m_{\rmn{e}} c^2 \beta_{\rmn{e}} \sqrt{\gamma-1}}{\hbar\,\omega_\rmn{plasma}}\right)\right.\nonumber \\
    &-&\ln(2)\left(\frac{\beta_{\rmn{e}}^2}{2}+\frac{1}{\gamma}\right)+\frac{1}{2}+
    \left.\left(\frac{\gamma-1}{4\gamma}\right)^2\right]\,, \\
  \left|\frac{d p}{d t}\right|_{\rm rad~} &=& \frac{4}{3}\frac{\sigma_\rmn{T}}{m_{\rmn{e}}c}\,
  \frac{p^2}{\beta_{\rmn{e}}}\,\left[1+\left(\frac{B}{B_\rmn{CMB}}\right)^2\right]\,.
\end{eqnarray}
Here $\beta_{\rmn{e}} = p/\sqrt{1+p^2}$ is the dimensionless velocity of CRs,
$\gamma=\sqrt{1+p^2}$ is the Lorentz factor of CRs,
$\omega_\rmn{plasma} = \sqrt{4\upi e^2 n_{\rmn{e}} / m_{\rmn{e}}}$ is the plasma
frequency, $n_e$ is the number density of free electrons, and
$\sigma_\rmn{T}= 8\upi e^4/3(m_{\rmn{e}} c^2)^2$ is the Thomson cross
section. The {\it rms} magnetic field strength is denoted by $B$ and
the equivalent field strength of the cosmic-microwave background is
given by $B_\rmn{CMB} = 3.24 (1 + z)^2\umu\rmn{G}$, where $z$ denotes
the redshift. In the peripheral cluster regions, where $B \ll
B_\rmn{CMB}$, the CRes loose virtually all their energy by means of
inverse Compton emission. $D_{pp}$ is the momentum space diffusion
coefficient (see \S\ref{sec:reacc}), and
$Q_{\rmn{e}}$ denotes the injection rate of primary and secondary
electrons in the ICM (see Sect.~\ref{sec:cosmo_sim}). The first term
containing the expression $\bnabla\bcdot \bvel$ represents Fermi-I
acceleration and the second term of this form describes adiabatic
gains and losses.

During post-processing of our Coma-like cluster simulation, we solve the
Fokker-Planck equation over a redshift interval from $z=5$ to 0. The simulated
cluster undergoes a major merger over the last 1-2~Gyrs that is thought to
inject large turbulent eddies. As is commonly assumed
\citep{brunetti07,brunetti11,2004ApJ...614..757Y,2013ApJ...771..131B} we assume
that about one Gyr after core passage the fields have decayed down to the
smallest scales $k_{\rm cut}$, and the radio halo turns on shortly after. We
choose this simulation snapshot to analyse. We are not very sensitive to this
assumption, since the thermal and CR quantities are very similar a few 100 Myrs
before and after $z = 0$, where we have chosen to evaluate the simulations. In
all our calculations we assume that turbulent reacceleration efficiently
accelerates particles for $\tau_{\rm cl} \sim 650$ Myrs (which is roughly the
cascade time on which turbulence is damped) and that during this turbulent phase
CR streaming and spatial diffusion can be neglected. In
\S\ref{sect:self-limiting}, we explore sensitivity to the last assumption.

Thus far, we have ignored CR transport. However, if CRps stream in the ICM, then
their spatial profile could potentially flatten sufficiently
\citep{ensslin11,wiener13}. This scenario is very attractive: it generates seed
electrons with the right spatial footprint, and by removing CRps from the core,
obeys gamma-ray constraints. Turbulence plays two opposing roles: Alfv{\'e}nic
turbulence damps waves generated by the CR streaming instability
\citep{yan02,farmer04}, thus reducing self-confinement; but compressible fast
modes scatter CRs directly. Turbulent damping is still efficient for highly
subsonic conditions \citep{wiener13}, while we assume compressible fast modes to
only provide effective spatial confinement during the periods of transsonic,
highly super-Alfv{\'e}nic ($\mathcal{M}_{\rm A} \sim 5$) turbulence associated with
mergers. Thus, CRs can stream out when the cluster is kinematically
quiescent. Furthermore, even Alfv{\'e}nic streaming timescales are relatively
short \cite[$\sim 0.1-0.5$ Gyr;][]{wiener13} compared to the timescale on which
the CRp population is built up. Based on these findings, we adopt a toy model
for our \Mstream scenario in which CR streaming instantaneously produces flat
CRp profiles. We assume that CRs cannot stream significantly past perpendicular
$B$-fields at the accretion shock, so that the total number of CRs is conserved
within the virial radius during the streaming process.  In all our calculations
we assume that turbulent reacceleration efficiently accelerates particles for
$\tau_\rmn{cl}\sim650$~Myrs (which is roughly the cascade time on which
turbulence is damped) and that during this turbulent phase CR streaming and
spatial diffusion can be neglected.

The time evolution of the spectral energy distribution of CRps,
$f_{\rmn{p}}$, is similarly given by:
\begin{eqnarray}
\lefteqn{
  {{d f_{\rmn{p}}(p,t)}\over{d t}} =
  {{\partial }\over{\partial p}}
  \left\{
  f_{\rmn{p}}(p,t)\left[ \left|{{d p}\over{d t}}\right|_{\rm Coul}
    + \frac{p}{3}\bnabla\bcdot \left(\bvel+\bvel_{\rmn{st}}\right)\right.\right.}
\nonumber\\
&&-\left.\left. {1\over{p^2}}{{\partial }\over{\partial p}}\left(p^2 D_{pp}\right)
\right]\right\} - f_{\rmn{p}}(p,t) \bnabla\bcdot \left(\bvel+\bvel_{\rmn{st}}\right) 
- \bvel_{\rmn{st}}\bcdot\bnabla f_{\rmn{p}}(p,t)
\nonumber\\
&&+\,\,\, {{\partial^2 }\over{\partial p^2}}
\left[ D_{pp} f_{\rmn{p}}(p,t) \right] - {{f_{\rmn{p}}(p,t)}\over{\tau_{\rm had}(p)}}
+ Q_{\rmn{p}}(p,t)\, ,
\label{eq:FP_p}
\end{eqnarray}
where $\bvel_{\rmn{st}}=-\varv_{\rmn{A}}\bnabla f_{\rmn{p}}/|\bnabla
f_{\rmn{p}}|$ is the streaming velocity in the isotropic transport
approximation, $\varv_{\rmn{A}}$ is the Alfv\'en speed,
and $Q_{\rmn{p}}(p,t)$ denotes the injection rate of shock accelerated
CRps as a function of momentum $p$ and time $t$ (see
Sect.~\ref{sec:cosmo_sim}). The timescale of
hadronic losses that produce pions via CRp collisions with thermal
protons of the ICM is given by 
\begin{eqnarray}
  \tau_{\rm had} = \left[c\,n_\rmn{th}\,\sigma^{+/-,0}(p)\right]^{-1}\,,
\end{eqnarray}
where we use the cross-section, $\sigma^{+/-,0}(p)$, given by the
fitting formula in \citet{1986ApJ...307...47D}.

\subsection{Turbulent reacceleration}
\label{sec:reacc}

In turbulent reacceleration, particles gain energy via Fermi II acceleration. Wave-particle energy exchange takes place via transit time damping \citep[TTD,][]{brunetti07,brunetti11}. The
TTD resonance requires the wave frequency to obey the resonance
condition, $\omega=k_\parallel\varv_\parallel$, where $k_\parallel$
and $\varv_\parallel$ are the parallel (projected along the magnetic
field) wavenumber of the compressible mode and particle velocity, respectively. This implies
that the particle transit time across the confining wave region
matches the wave period,
$\lambda_{\parallel}/\varv_{\parallel}=\tau_{\rmn{wave}}$ (hence the moniker, 'transit time damping'). Note that the CRs'
gyroradius does not enter the resonance condition. Hence the CRs that
are in resonance with compressible waves experience Fermi-II
acceleration irrespective of the length scale of the perturbation.

However, the resonance changes the component of the particle momentum
parallel to the mean magnetic field, which over time leads to
increasing anisotropy in the particle distribution that decreases the
efficiency of reacceleration with time. As in \citet{brunetti11}, we
assume that there exists a mechanism---such as the firehose
instability---that isotropises the CR distribution function at the
gyroscale and on the reacceleration time scale, which ensures
sustained efficient reacceleration with time.
